Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-12 paper-new/appendix.tex
small changes to appendix

  % https://q.uiver.app/#q=WzAsNixbMCwwLCJBIl0sWzEsMCwiQSJdLFsyLDAsIkEnIl0sWzAsMSwiQSJdLFsxLDEsIkEnIl0sWzIsMSwiQSciXSxbMCwxLCJyKEEpIl0sWzEsMiwiYyJdLFszLDQsImMnIiwyXSxbNCw1LCJyKEEnKSIsMl0sWzAsMywiaV9BKFxcZGVsbGVfe2MnfSkiLDJdLFsxLDQsImYiLDJdLFsyLDUsImlfe0EnfShcXGRlbHJlX2MpIl1d
  \[\begin{tikzcd}[ampersand replacement=\&]
    A \& A \& {A'} \\
    A \& {A'} \& {A'}
    \arrow["{r(A)}", from=1-1, to=1-2]
    \arrow["{i_A(\delle_{c'})}"', from=1-1, to=2-1]
    \arrow["c", from=1-2, to=1-3]
    \arrow["f"', from=1-2, to=2-2]
    \arrow["{i_{A'}(\delre_c)}", from=1-3, to=2-3]
    \arrow["{c'}"', from=2-1, to=2-2]
    \arrow["{r(A')}"', from=2-2, to=2-3]
  \end{tikzcd}\]

@@

    % https://q.uiver.app/#q=WzAsOSxbMCwwLCJcXGxpIEEnJyJdLFsyLDAsIlxcbGkgQScnIl0sWzAsMSwiXFxsaSBBJyciXSxbMiwxLCJcXGxpIEEnJyJdLFswLDIsIlxcbGkgQSJdLFsyLDIsIlxcbGkgQScnIl0sWzAsMywiXFxsaSBBIl0sWzIsMywiXFxsaSBBJyciXSxbMSwyLCJcXGxpIEEnIl0sWzAsMSwicihcXGxpIEEnJykiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsyLDMsInIoXFxsaSBBJycpIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNiw3LCJcXGxpIChjIFxcY29tcCBjJykiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDIsImlfe0EnJ30oXFxkZWx0YV5yKSIsMl0sWzIsNCwicF97RmMgXFxjb21wIEZjJ30iLDJdLFs0LDYsIlxcaWQiLDJdLFsxLDMsImlfe0EnJ30oXFxkZWx0YV5yKSJdLFszLDUsImlfe0EnJ30oXFxkZWxycF97RmMgXFxjb21wIEZjJ30pIl0sWzUsNywiXFxpZCJdLFs0LDgsIlxcbGkgYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzgsNSwiXFxsaSBjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEyLDE1LCJcXGlkIiwzLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEzLDE2LCJcXGRubF97XFxsaSBjIFxcY29tcCBcXGxpIGMnfSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
    \[\begin{tikzcd}[ampersand replacement=\&]
      {\li A''} \&\& {\li A''} \\
      {\li A''} \&\& {\li A''} \\
      {\li A} \& {\li A'} \& {\li A''} \\
      {\li A} \&\& {\li A''}
      \arrow["{r(\li A'')}", "\shortmid"{marking}, no head, from=1-1, to=1-3]
      \arrow[""{name=0, anchor=center, inner sep=0}, "{i_{A''}(\delta^r)}"', from=1-1, to=2-1]
      \arrow[""{name=1, anchor=center, inner sep=0}, "{i_{A''}(\delta^r)}", from=1-3, to=2-3]
      \arrow["{r(\li A'')}", "\shortmid"{marking}, no head, from=2-1, to=2-3]
      \arrow[""{name=2, anchor=center, inner sep=0}, "{p_{Fc \comp Fc'}}"', from=2-1, to=3-1]
      \arrow[""{name=3, anchor=center, inner sep=0}, "{i_{A''}(\delrp_{Fc \comp Fc'})}", from=2-3, to=3-3]
      \arrow["{\li c}", "\shortmid"{marking}, no head, from=3-1, to=3-2]
      \arrow["\id"', from=3-1, to=4-1]
      \arrow["{\li c'}", "\shortmid"{marking}, no head, from=3-2, to=3-3]
      \arrow["\id", from=3-3, to=4-3]
      \arrow["{\li (c \comp c')}"', "\shortmid"{marking}, no head, from=4-1, to=4-3]
      \arrow["\id"{marking, allow upside down}, draw=none, from=0, to=1]
      \arrow["{\dnl_{\li c \comp \li c'}}"{description}, draw=none, from=2, to=3]
    \end{tikzcd}\]

@@

\eric{Left off here}

@@

\begin{lemma}
  The following hold:
  \begin{itemize}
    \item $U(d \comp d') \qordeq U(d) \comp U(d')$
    \item $\li(c \comp c') \qordeq \li (c) \comp \li (c')$
    \item $(c \comp c') \to (d \comp d') \qordeq (c \to d) \comp (c' \to d')$
    \item $(c_1 \comp c_1') \times (c_2 \comp c_2') \qordeq (c_1 \times c_2) \comp (c_1'\times c_2')$
  \end{itemize}
\end{lemma}

@@

>>>>

  % https://q.uiver.app/#q=WzAsNixbMCwwLCJBIl0sWzEsMCwiQSJdLFsyLDAsIkEnIl0sWzAsMSwiQSJdLFsxLDEsIkEnIl0sWzIsMSwiQSciXSxbMCwxLCJyKEEpIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMyw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzQsNSwicihBJykiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDMsImlfQShcXGRlbGxlX3tjJ30pIiwyXSxbMSw0LCJmIiwyXSxbMiw1LCJpX3tBJ30oXFxkZWxyZV9jKSJdXQ==
  \[\begin{tikzcd}[ampersand replacement=\&]
    A \& A \& {A'} \\
    A \& {A'} \& {A'}
    \arrow["{r(A)}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
    \arrow["{i_A(\delle_{c'})}"', from=1-1, to=2-1]
    \arrow["c", "\shortmid"{marking}, no head, from=1-2, to=1-3]
    \arrow["f"', from=1-2, to=2-2]
    \arrow["{i_{A'}(\delre_c)}", from=1-3, to=2-3]
    \arrow["{c'}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
    \arrow["{r(A')}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
  \end{tikzcd}\]

@@

    % https://q.uiver.app/#q=WzAsOSxbMCwwLCJcXGxpIEEnJyJdLFsyLDAsIlxcbGkgQScnIl0sWzAsMSwiXFxsaSBBJyciXSxbMiwxLCJcXGxpIEEnJyJdLFswLDIsIlxcbGkgQSJdLFsyLDIsIlxcbGkgQScnIl0sWzAsMywiXFxsaSBBIl0sWzIsMywiXFxsaSBBJyciXSxbMSwyLCJcXGxpIEEnIl0sWzAsMSwicihcXGxpIEEnJykiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsyLDMsInIoXFxsaSBBJycpIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNiw3LCJcXGxpIChjIFxcY29tcCBjJykiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDIsImlfe0EnJ30oXFxkZWx0YV5yKSIsMl0sWzIsNCwicF97XFxsaSBjIFxcY29tcCBcXGxpIGMnfSIsMl0sWzQsNiwiXFxpZCIsMl0sWzEsMywiaV97QScnfShcXGRlbHRhXnIpIl0sWzMsNSwiaV97QScnfShcXGRlbHJwX3tcXGxpIGMgXFxjb21wIFxcbGkgYyd9KSJdLFs1LDcsIlxcaWQiXSxbNCw4LCJcXGxpIGMiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs4LDUsIlxcbGkgYyciLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxMiwxNSwiXFxpZCIsMyx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxMywxNiwiXFxkbmxfe1xcbGkgYyBcXGNvbXAgXFxsaSBjJ30iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
    \[\begin{tikzcd}[ampersand replacement=\&]
      {\li A''} \&\& {\li A''} \\
      {\li A''} \&\& {\li A''} \\
      {\li A} \& {\li A'} \& {\li A''} \\
      {\li A} \&\& {\li A''}
      \arrow["{r(\li A'')}", "\shortmid"{marking}, no head, from=1-1, to=1-3]
      \arrow[""{name=0, anchor=center, inner sep=0}, "{i_{A''}(\delta^r)}"', from=1-1, to=2-1]
      \arrow[""{name=1, anchor=center, inner sep=0}, "{i_{A''}(\delta^r)}", from=1-3, to=2-3]
      \arrow["{r(\li A'')}", "\shortmid"{marking}, no head, from=2-1, to=2-3]
      \arrow[""{name=2, anchor=center, inner sep=0}, "{p_{\li c \comp \li c'}}"', from=2-1, to=3-1]
      \arrow[""{name=3, anchor=center, inner sep=0}, "{i_{A''}(\delrp_{\li c \comp \li c'})}", from=2-3, to=3-3]
      \arrow["{\li c}", "\shortmid"{marking}, no head, from=3-1, to=3-2]
      \arrow["\id"', from=3-1, to=4-1]
      \arrow["{\li c'}", "\shortmid"{marking}, no head, from=3-2, to=3-3]
      \arrow["\id", from=3-3, to=4-3]
      \arrow["{\li (c \comp c')}"', "\shortmid"{marking}, no head, from=4-1, to=4-3]
      \arrow["\id"{marking, allow upside down}, draw=none, from=0, to=1]
      \arrow["{\dnl_{\li c \comp \li c'}}"{description}, draw=none, from=2, to=3]
    \end{tikzcd}\]

@@

@@

\begin{lemma}\label{lem:quasi-order-equiv-functors}
  The following hold:
  \begin{itemize}
    \item $U(d \comp d') \qordeq U(d) \comp U(d')$
    \item $\li(c \comp c') \qordeq \li (c) \comp \li (c')$
    \item $(c \comp c') \to (d \comp d') \qordeq (c \to d) \comp (c' \to d')$
    \item $(c_1 \comp c_1') \times (c_2 \comp c_2') \qordeq (c_1 \times c_2) \comp (c_1'\times c_2')$
  \end{itemize}
\end{lemma}
\begin{proof}
  (1) and (2) were already shown in Lemmas \ref{lem:Fcc-equiv-FcFc'} and
  \ref{lem:Udd-equiv-UdUd'}. (3) follows from observing that both relations are
  right-represented by the morphism $e_c'e_c \arr p_{d}p_{d'}$ and (4) from the
  fact that both are left-represented by the morphism $e_{c_1'} e_{c_1} \times
  e_{c_2'} e_{c_2}$
\end{proof}

@@

\subsection{Definitions of Error Ordering and Weak Bisimilarity for the Delay Monad}
\label{sec:relations-on-delay-monad}

Correspondingly, we define a notion of lock-step error ordering and weak
bisimilarity relation for the coinductive delay monad $\delay(\mathbb{N} + {\mho})$:

The lock-step error ordering is defined coinductively by the following rules:

%
\begin{mathpar}
  \inferrule*[]
  { }
  {\tnow (\inr\, 1) \ledelay d}

  \inferrule*[]
  {x_1 \le_X x_2}
  {\tnow (\inl\, x_1) \ledelay \tnow (\inl\, x_2)}

  \inferrule*[]
  {d_1 \ledelay d_2}
  {\tlater\, d_1 \ledelay \tlater\, d_2}
\end{mathpar}
%
And we similarly define by coinduction a ``weak bisimilarity'' relation on
$\delay(\mathbb{N} + {\mho})$. This uses a relation $d \Da x_?$ between
$\delay(\mathbb{N} + {\mho})$ and $\mathbb{N} + {\mho}$ that is defined as $d
\Da n_? := \Sigma_{i \in \mathbb{N}} d = \tlater^i(\tnow\, n_?)$. Then weak
bisimilarity for the delay monad is defined coinductively by the rules
%
\begin{mathpar}
  \inferrule*[]
  {n_? \bisim_{\mathbb{N} + {\mho}} m_?}
  {\tnow\, n_? \bisimdelay \tnow\, m_? }

  \inferrule*[leftskip=1.5em]
  {d_1 \Da n_? \and n_? \bisim_{\mathbb{N} + {\mho}} m_?}
  {\tlater\, d_1 \bisimdelay \tnow\, m_? }

  \inferrule*[leftskip=1.5em]
  {d_2 \Da m_? \and n_? \bisim_{\mathbb{N} + {\mho}} m_?}
  {\tnow\, n_? \bisimdelay \tlater\, d_2}

  \inferrule*[leftskip=1.5em]
  {d_1 \bisimdelay d_2}
  {\tlater\, d_1 \bisimdelay \tlater\, d_2 }
  %
  % \inferrule*[]
  % {d_1 \Da x_? \and d_2 \Da y_? \and x_? \bisim_{X + 1} y_?}
  % {d_1 \bisimdelay d_2}
  %
  % \inferrule*[]
  % {d_1 \bisimdelay d_2}
  % {\tlater d_1 \bisimdelay \tlater d_2 }
\end{mathpar}
